Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex
fix bib ref in between collumns
foo

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\LLVMMac{LLVM-Mac\xspace}
\def\GCCLinux{GCC-Linux\xspace}
\def\plotHeight{7cm}

@@

   \end{subfigure}
   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. Higher than 1 is better.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.047 $\pm$ 0.0088 & 23.492 $\pm$ 0.0135 \\
  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.6238 $\pm$ 0.0012 & 1.5807 $\pm$ 0.0092 & 11.4096 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
   \hline
  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.241 $\pm$ 0.0515 & 22.542 $\pm$ 0.0775\\
 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.6900 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.149 $\pm$ 0.0321  & 12.939 $\pm$ 0.0177 \\
   \hline
  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.657 $\pm$ 0.0481 & 24.927 $\pm$ 0.0309\\
 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.643 $\pm$ 0.06  & 24.881 $\pm$ 0.1983 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

\begin{figure*}[b!]
		\centering
		\makebox[0pt]{
		\includegraphics[width=1.09\linewidth]{figures/asm_cmp.pdf}}
		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
		\label{fig:Asm}
\end{figure*}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

@@

\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when LLVM is used, with respectively 12,94 and 24,88 relative speed-up factors.

This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors. On the contrary, \LLVMMac generates more instructions: it especially fails at preserving registers holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.

@@

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation. The results showed the \legRTL implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.

>>>>

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\LLVMMac{LLVM-Mac\xspace}
\def\GCCLinux{GCC-Linux\xspace}
\def\plotHeight{6.8cm}

@@

   \end{subfigure}
   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. The higher the better.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 23.49 $\pm$ 0.0135 \\
  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.624 $\pm$ 0.0012 & 1.581 $\pm$ 0.0092 & 11.41 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
   \hline
  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 22.54 $\pm$ 0.0775\\
 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.690 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.15 $\pm$ 0.0321  & 12.94 $\pm$ 0.0177 \\
   \hline
  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 24.93 $\pm$ 0.0309\\
 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.0600  & 24.88 $\pm$ 0.1983 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

\begin{figure*}[b!]
		\centering
		\makebox[0pt]{
		\includegraphics[width=\linewidth]{figures/asm_cmp.pdf}}
		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
		\label{fig:Asm}
\end{figure*}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings. The problem is even more important for the plugin version, since the beginning of the primitive code, before the main comparison loop, is slower due to the calls to the main VM code.
As the string size goes larger, the overhead is less significant because the execution time becomes dominated by the time spent in the main comparison loop.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

@@

\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.

This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version. As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar. On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in the figure).

This is an example of one native function in the whole VM and we do not conclude anything about the C compiler capabilites from it. VMs compiled with LLVM and GCC are around equally fast except in narrow cases like this one and LLVM is able to optimize correctly the \legSlangPlug version. LLVM likely fails at optimizing the \legSlang version likely due to the old C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, historically the only compiler we supported. However, we can see that the \legRTL version has the most reliable performance since it does not depend on what optimization the C compiler is able to perform or not. This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support, where C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into such optimizations.

@@

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation. The results showed the \legRTL implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack switching, optimization for large strings, and the control over generated code.
